CUBE: An Information-optimized parallel Cosmological $N$-body Algorithm

### 2018-03-16 ms_caf/ms_caf.tex
reply referee of CUBE paper

When $N$ is large, the brute force pairwise particle-particle (PP) force brings unaffordable $o(N^2)$ computations. However the workload can be alleviated to $o(N\log N)$ by many algorithms. For example, the ``tree'' algorithms like {\tt GADGET} \citep{2001NewA....6...79S,2005MNRAS.364.1105S} and its simulation ``Millennium'' \citep{2005Natur.435..629S,2012MNRAS.426.2046A}, {\tt TPM} \citep{1995ApJS...98..355X} and {\tt GOTPM} \citep{2004NewA....9..111D}, adaptive grid algorithm including {\tt HYDRA} \citep{1995ApJ...452..797C} and {\tt RAMSES} \citep{2010ascl.soft11007T}, as well as mesh-refined codes \citep{1991ApJ...368L..23C} and moving adaptive particle-mesh (PM) codes \citep{1995ApJS..100..269P}. Among these, the standard PM algorithm \citep{1988csup.book.....H} is most memory and computational efficient if we focus on large cosmological scales. The load balancing problem is minimized as the matter distribution is rather homogeneous, and the speed benefits from the fast Fourier transform (FFT) libraries, such as {\tt FFTW3} \citep{Frigo05thedesign}. {\tt PMFAST} \citep{2005NewA...10..393M} is first codes to use a 2-level PM algorithm, aiming to put PM codes toward speed, memory compactness and scalability. After subsequent developments on {\tt PMFAST}, {\tt CUBEP3M} \citep{2013MNRAS.436..540H} uses cubic spatial decomposition, and adds PP force and many other features.

@@

In the following sections we present a new, information-optimized parallel cosmological $N$-body simulation code {\tt CUBE}, using as low as 6 byte per particle. It gives accurate results in cosmological LSS simulations -- the error induced by information optimization is below the error from the PM algorithm. In \S\ref{s.method} we show how the memory can be saved by using a ``integer-based storage'' and how the PM $N$-body algorithm is adapted with this storage format. In \S\ref{s.results} we quantify the accuracy of this algorithm by groups of simulations by {\tt CUBEP3M} and {\tt CUBE}. Discussions and conclusions are in \S\ref{s.discussion}.

@@

\subsection{Particle position storage}\label{ss.position}
We construct a uniform mesh throughout the space and each particle belongs to its parent cell of the mesh. Instead of storing global coordinates of each particle, we store its offset relative to its parent cell which contains the particle. We divide the cell, in each dimension $d$, evenly into $2^8=256$ bins, and use a 1-byte (8 bits) integer $\chi_d \in \{-128,-127,...,127\}$ to indicate which bin it locates in this dimension. The global locations of particles are given by cell-ordered format in memory space, and a complimentary number count of particle number in this mesh (density field) will give complete information of particle distribution in the mesh. Then the global coordinate in $d$th dimension $x_d$ is given by $x_d=(n_c-1)+(\chi_d+128+1/2)/256$, where $n_c=1,2,...,N_c$ is the index of the coarse grid. The mesh is chosen to be coarse enough such that the density field takes negligible memory. This coarse density field can be further compressed into 1-byte integer format, such that a 1-byte integer show the particle number in this coarse cell in range 0 to 255. In the densest cells (rarely happened) where there are $\ge 255$ particles, we can just write 255, and write the actual number as a 4-byte integer in another file.

In a simulation with volume $L^3$ and $N_c^3$ coarse cells, particle positions are stored with a precision of $L/(256N_c)$. The force calculation (e.g. softening length) should be configured much finer than this resolution, discussed in later sections. On the other hand, particle position can also be stored as 2-byte (16 bits) integers to increase the resolution. In this case, each coarse cell is divided into $2^{16}=65536$ bins and the position precision is $L/(65536N_c)$, precise enough compared to using4-byte global coordinates, see later results. We denote this case ``x2'' and denote using 1-byte integers for positions ``x1''.

We collectively write the general position conversion formulae
\begin{equation}\label{eq.chi}
	\chi_d=\left[2^{8n_\chi}(x_d-\left[x_d\right])\right]-2^{8n_\chi-1},
\end{equation}
\begin{equation}\label{eq.x}
	x_d=(n_c-1)+2^{-8n_\chi}\left(\chi_d+2^{8n_\chi-1}+1/2\right),
\end{equation}
where $[\ ]$ is the operator to take the integer part. $n_\chi\in\{1,2\}$ is the number of bytes used for each integer, $x_d$ and $\chi_d$ are floating and integer version of the coordinate. The velocity counterpart of them are $n_\nu=1,2$, $v_d$ and $\nu_d$. $n_c=1,2,...$ is the coarse grid index, given by the ordering of particles and a integer based particle density field (see \ref{sss.spatial_decomposition}). The position resolution for $n_\chi$-byte integer, ``x$n_\chi$'', is $2^{-8n_\chi}L/N_c$.

@@

\subsection{Particle velocity storage}\label{ss.velocity}
Similarly, actual velocity in $d$th dimension $v_d$ is decomposed into an averaged velocity field on the same coarse grid $v_c$ and a residual $\Delta v$ relative to this field: 
\begin{equation}
	v_d=v_c+\Delta v.
\end{equation}
$v_c$ is always recorded and kept updated, and occupies negligible memory. We then divide velocity space $\Delta v$ into uneven bins, and use a $n_\nu$-byte integer to indicate which $\Delta v$ bin the particle is located.

@@

\begin{figure}
\centering
  \includegraphics[width=0.9\linewidth]{f2}
 \caption{Spacial decomposition in {\tt CUBE} in a 2D analogy. In this example, there are 2 images per dimension (${\tt nn}=2$), and 2 tiles per image per dimension (${\tt nnt}=2$). The orange boxes show the overlapped physical+buffer=extended tile regions, inside of which one physical regions is indicated with green.}
\label{fig.tile}
\end{figure}

The simulation results are very insensitive if we manually tune the variance of the model $\sigma_\Delta$ within an order of magnitude. However, in $n_\nu=1$ case, the method of using uneven bins gets much better results than simply using equal bins between minimum and maximum values $[\min(\Delta v),\max(\Delta v)]$. So, one can safely use a standard $\Lambda$CDM (cold dark matter with a cosmological constant as dark energy) for slightly different cosmological models, in equation (\ref{eq.vdisp}). In {\tt CUBE}, the velocity conversion takes the formula
\begin{equation}\label{eq.nu}
	\nu_d=\left\lfloor(2^{8n_\nu}-1)\pi^{-1}\tan^{-1}\left((v_d-v_c)\sqrt{\pi/2\sigma_\Delta^2}\right)\right\rfloor,
\end{equation}
\begin{equation}\label{eq.v}
	v_d=v_c+\tan\left(\frac{\pi\nu_d}{2^{8n_\nu}-1}\right)\sqrt{2\sigma_\Delta^2/\pi},
\end{equation}
where $\lfloor\ \rfloor$ is the operator to take the nearest integer. Tangent functions are convenient and compute very fast. Compared to error functions used in Gaussian case, they take the same variance at $\nu_d=0$ but resolve high velocities relatively better. Note again that proper choice of conversion formulae and $\sigma_\Delta$ only optimizes the velocity space sampling, but does not affect the physics.

@@

\subsection{Code overview}\label{ss.overview}
{\tt CUBE} uses a 2-level PM force calculation, same as {\tt CUBEP3M}, However, in order to apply the integer based format to the $N$-body simulation, substantial structural changes need to be done. {\tt CUBE} is written in Coarray Fortran, where Coarray features replace MPI (Message Passing Interface) communications between computation nodes/images\footnote{Images are the concept of computing nodes or MPI tasks in Coarray Fortran. We use this terminology in this paper.}. The algorithm is described in this language.

\subsubsection{Spacial decomposition}\label{sss.spatial_decomposition}
{\tt CUBE} uses cubic decomposition structures. The global simulation volume is decomposed into ${\tt nn}^3$ cubic sub-volumes with {\tt nc} coarse grids per side, and each of these are assigned to a coarray {\it image}. Inside of an image, the sub-volume is further decomposed into ${\tt nnt}^3$ cubic {\it tiles}, with {\tt nt}={\tt nc}/{\tt nnt} coarse grids per side, which is a essential unit in the calculations. Each tile is surrounded by a {\it buffer} region which is ${\tt ncb}$ coarse cells thick. The buffer is designed for two reasons: (1) computing the fine mesh force, whose cut-off length ${\tt nforce\_cutoff}\leq{\tt ncb}$, and (2) collecting all possible particles travelling from a tile's buffer region to its center, {\it physical} region. Figure \ref{fig.tile} shows the spacial decomposition in a 2-dimensional analogy, with ${\tt nn=2}$ and ${\tt nnt=2}$.

According to this spatial decomposition, and as discussed in the last two subsections, we declare $\{\rho_c,\chi_d,v_c,\nu_d\}$, or in variable expression $\{{\tt rho\_c},{\tt xp},{\tt vfield},{\tt vp}\}$, as follows:
\begin{eqnarray*}
	&&{\tt integer(1)\ rho\_c(nex,nex,nex,nnt,nnt,nnt)[nn,nn,*]}\\
	&&{\tt integer(}n_\chi{\tt )\ xp(npmax)[nn,nn,*]}\\
	&&{\tt real(4)\ vfield(nex,nex,nex,nnt,nnt,nnt)[nn,nn,*]}\\
	&&{\tt integer(}n_\nu{\tt )\ vp(npmax)[nn,nn,*]}
\end{eqnarray*}
where ${\tt nex}={\tt nt}+2\times{\tt ncb}$ covers the buffer region on both sides, {\tt nnt} is the tile dimemsions, and {\tt nn} is the image {\it codimensions}\footnote{Coarray Fortran concept. Codimensions can do communications between images.}. We denote actual number of particles in a given image {\tt nplocal}, and ${\tt npmax}>{\tt nplocal}$ is a value (discussed in Section \ref{ss.memory}) large enough to store particles. The particle position and velocity arrays {\tt xp} and {\tt vp} (equivalent to $\chi_d$ and $\nu_d$ in \ref{ss.position} and \ref{ss.velocity} respectively) are required to be sorted according to the same memory layout of {\tt rho\_c}, such that $n_c$ and thus global positions of particles $x_d$ can be obtained by equation (\ref{eq.x}). $\{{\tt rho\_c},{\tt xp},{\tt vfield},{\tt vp}\}$ provides a complete information of positions and velocities of particles, and we call it a snapshot, or checkpoint.

An additional particle-ID (PID) array, {\tt PID}, can also be declared to differentiate particle types (e.g. CDM and neutrino particles) or differentiate every particle, by using 1-byte, 4-byte or 8-byte integer per particle. If PID is turned on, the array {\tt PID} is also included in the checkpoints, and the ordering of PID is same as {\tt xp} and {\tt vp}.

\begin{figure}[t]
\begin{verbatim}

program Initial_Condition_Generator_for_CUBE
  Get random seed
  Generate Gaussian random field
  Multiply by power spectrum in Fourier space
  Get linear density field
  Solve Poisson equation
  Get displacement potential  
  do (each tile) ! create particles per tile
    do (each fine grid in the extended tile region)
      calculate particle's coarse grid position
      update rho_c_new
    enddo
    do (each fine grid in the extended tile region)
      calculate particle's new accurate position
      calculate praticle's initial velocity
      calculate particle's index
      create xp_new(index) & vp_new(index)
      if (PID_flag) create PID(index)
    enddo
    do (each coarse grid)
      discard buffer information
    enddo
    write to disk
    sum up nplocal
  enddo
  sync all
  sum up npglobal
  write to headers of the checkpoint
end
\end{verbatim}
\caption{Pseudocode for the initial condition generator.}
\label{fig.ic}
\end{figure}

@@

The first step is the calculation of the potential field. At a sufficient high initial redshift $z_i$, we first generate a linear density fluctuation $\delta_L({\bs q})$ on Lagrangian grid ${\bs q}$ by multiplying a Gaussian random field with the transfer function $T(k)$ in Fourier space, where $T(k)$ is given by the assumed cosmological model. Then we solve for the potential ${\Phi({\bs q}, z_i)}$ of a curl-less displacement field ${\bs\Psi({\bs q})}$ by Poisson equation $-\nabla^2\Phi=\delta_L$. At this step, global Pencil FFTs are applied on the fine mesh to transform the initial random Gaussian field to the displacement potential $\Phi$. The initial condition generator is applied only once, and in principle, at some costs of speed, one can use only one scalar field on the fine mesh, and FFTs can be done in-place. If the number of particles equals the number of fine grids, then the memory consumption is dominated by the single precision fine mesh field, or 4 bpp.

The second step is to generate particles on Lagrangian coordinates ${\bs q}$ and displace them with Zel'dovich approximation (ZA) \citep{1970A&A.....5...84Z}, by locally differentiating $\Phi$ in real space. This step is done tile by tile. For each tile, we iterate twice over particles' Lagrangian coordinates ${\bs q}$ (usually at the center of every fine grid). The first iteration calculates particles' new, coarse grid positions by ZA and get a coarse density field on tile. The second iteration, ZA is applied again to calculate the accurate positions of particles and their velocities, and convert them into integer based ones by Equations (\ref{eq.chi},\ref{eq.nu}) and place them in a certain order according to the coarse density field generated by the first iteration. Lastly, we discard those particles out of the physical region of the tile, and write only particles in physical region on to the disk. As we will see in Section \ref{sss.algorithm}, {\tt update\_xp} paragraph, it is very similar to the ``drift'' step in the main $N$-body code, where an ``buffered'' state of particles is firstly created by two iterations of particles, and the particles in the buffer region are discarded to achieve a ``disjoint'' state. If PIDs are needed, they are also generated at the second iteration and are manipulated similarly as integer based velocities. Because particles are generated and saved to disk tile by tile, locally, after working on certain tile, the arrays for positions, velocities and PIDs (if any) are freed up and can be reused for the next tile. During this step, the only major memory usage is still dominated by the fine grid potential $\Phi$.

After working on all tiles, we sum up the total particle number on this image, {\tt nplocal}, and write it in the header of the checkpoint. Finally, $\{{\tt rho\_c},{\tt xp},{\tt vfield},{\tt vp},{\tt PID}({\rm optional})\}$ is rewritten to the disk as the checkpoint of $z=z_i$. We summarize the above steps into a pseudocode in Figure \ref{fig.ic}, where the ``{\tt do}'' loop to create particles per tile is very similar to the following {\tt update\_xp} subroutine in the main $N$-body code. The corresponding arrays are also discussed below.

@@

\subsubsection{Algorithm}\label{sss.algorithm}
Figure \ref{fig.code} shows the overall structure of the code and these subroutines are described in following paragraphs.

{\tt initialize} and {\tt read\_particles} -- The subroutine {\tt initialize} creates necessary FFT plans and read in configuration files telling the program at which redshifts we need to do checkpoints, halofinds, or stop the simulation. Force kernels {\tt kern\_c}, {\tt kern\_f} are also created or read in here. In {\tt read\_particles}, for each image, we read in all particles in {\it physical} regions of every tile (indicated in Figure \ref{fig.tile}), i.e., $\{{\tt rho\_c},{\tt xp},{\tt vfield},{\tt vp},{\tt PID}({\rm if\ any})\}$. These are obtained by the initial condition generator (Section \ref{sss.ic}). Because physical regions of tiles are {\it complete} and {\it disjoint} in space, particles at this stage are also complete and disjoint. We call it ``disjoint state''. At this stage, {\tt rho\_c}'s values in buffer regions are 0's, and the elements of {\tt xp} and {\tt vp} beyond physical number -- {\tt xp(nplocal+1:)} and {\tt vp(nplocal+1:)} can be arbitrary and are not used.

{\tt buffer\_density}, {\tt buffer\_x} and {\tt buffer\_v} -- In order to use the integer based format, {\tt xp} and {\tt vp} must always be ordered, and their number density field {\tt rho\_c} must always be present. In updating arrays of {\tt xp} and {\tt vp} (not simultaneously) of a local tile, the buffer region of the tile is also needed. First, buffer regions of {\tt rho\_c} is synchronized between tiles and images by subroutine {\tt buffer\_density}. Then, by subroutines {\tt buffer\_x} and {\tt buffer\_v} respectively, {\tt xp} and {\tt vp} are updated to contain common, buffered particles, in an order according to the new, buffered {\tt rho\_c}. If there are PIDs used, then {\tt PID} is also updated in subroutine {\tt update\_v}. We call this stage ``buffered state''. After {\tt particle\_initialize} is done by all images (synchronized by a ``{\tt sync all}'', which is equivalent to a {\tt mpi\_barrier}), these three subroutines are called and particles are converted from disjoint state to buffered state.

{\tt timestep} --
We operate a Runge-Kutta 2 method for time integration. i.e., updating positions ($D$=drift) and velocities ($K$=kick) are at every half time step by operator splitting. For $n$ time steps, the operation would be $(DKKD)^n$ which is 2nd order accurate. The actual simulation applies varied time steps. In each iteration of the main loop, we firstly call {\tt timestep}, where a increment of time {\tt dt} is controlled by particles' maximum velocities, accelerations, cosmic expansion and any other desired conditions.

>>>>

When $N$ is large, the brute force pairwise particle-particle (PP) force brings unaffordable $o(N^2)$ computations, so many algorithms were designed to alleviate it. Various fast multipole methods \citep{1985JCoPh..60..187R} improve the complexity to $o(N\log N)$ even $o(N)$, among which the most popular one is ``tree'', like {\tt GADGET} \citep{2001NewA....6...79S,2005MNRAS.364.1105S} and its simulation ``Millennium'' \citep{2005Natur.435..629S,2012MNRAS.426.2046A}, {\tt TPM} \citep{1995ApJS...98..355X} and {\tt GOTPM} \citep{2004NewA....9..111D}. Other methods include adaptive grid algorithms like {\tt HYDRA} \citep{1995ApJ...452..797C} and {\tt RAMSES} \citep{2010ascl.soft11007T}, as well as mesh-refined codes \citep{1991ApJ...368L..23C} and moving adaptive particle-mesh (PM) codes \citep{1995ApJS..100..269P}. The standard PM algorithm \citep{1988csup.book.....H} is most memory and computational efficient if we focus on large cosmological scales. The load balancing problem is minimized as the matter distribution is rather homogeneous, and the speed benefits from the fast Fourier transform (FFT) libraries, such as {\tt FFTW3} \citep{Frigo05thedesign}. {\tt PMFAST} \citep{2005NewA...10..393M} is first codes to use a 2-level PM algorithm, aiming to put PM codes toward speed, memory compactness and scalability. After subsequent developments on {\tt PMFAST}, {\tt CUBEP3M} \citep{2013MNRAS.436..540H} uses cubic spatial decomposition, and adds PP force and many other features.

@@

In the following sections we present a new, information-optimized parallel cosmological $N$-body simulation code {\tt CUBE} \citep{haoran_yu_2018_1198909}, using as low as 6 byte per particle. It gives accurate results in cosmological LSS simulations -- the error induced by information optimization is below the error from the PM algorithm. In \S\ref{s.method} we show how the memory can be saved by using a ``integer-based storage'' and how the PM $N$-body algorithm is adapted with this storage format. In \S\ref{s.results} we quantify the accuracy of this algorithm by groups of simulations by {\tt CUBEP3M} and {\tt CUBE}. Discussions and conclusions are in \S\ref{s.discussion}.

@@

\subsection{Particle position storage}\label{ss.position}
We construct a uniform mesh throughout the space and each particle belongs to its parent cell of the mesh. Instead of storing global coordinates of each particle, we store its offset relative to its parent cell which contains the particle. This is similar with storing the quantities of nodes/clumps (structures of a tree in a tree code) relative to their parents \citep{1985SJSSC...6...85A}. We divide the cell, in each dimension $d$, evenly into $2^8=256$ bins, and use a 1-byte (8 bits) integer $\chi_d \in \{-128,-127,...,127\}$ to indicate which bin it locates in this dimension. The global locations of particles are given by cell-ordered format in memory space, and a complimentary number count of particle number in this mesh (density field) will give complete information of particle distribution in the mesh. Then the global coordinate in $d$th dimension $x_d$ is given by $x_d=(n_c-1)+(\chi_d+128+1/2)/256$, where $n_c=1,2,...,N_c$ is the index of the coarse grid. The mesh is chosen to be coarse enough such that the density field takes negligible memory. This coarse density field can be further compressed into 1-byte integer format, such that a 1-byte integer show the particle number in this coarse cell in range 0 to 255. In the densest cells (rarely happened) where there are $\ge 255$ particles, we can just write 255, and write the actual number as a 4-byte integer in another file.

In a simulation with volume $L^3$ and $N_c^3$ coarse cells, particle positions are stored with a resolution of $L/(256N_c)$. The force calculation (e.g. softening length) should be configured much finer than this resolution, discussed in later sections. On the other hand, particle position can also be stored as 2-byte (16 bits) integers to increase the resolution. In this case, each coarse cell is divided into $2^{16}=65536$ bins and the position resolution is $L/(65536N_c)$, precise enough compared to using4-byte global coordinates, see later results. We denote this case ``x2'' and denote using 1-byte integers for positions ``x1''.

We collectively write the general position conversion formulae
\begin{equation}\label{eq.chi}
	\chi_d=\left[2^{8n_\chi}(x_d-\left[x_d\right])\right]-2^{8n_\chi-1},
\end{equation}
\begin{equation}\label{eq.x}
	x_d=(n_c-1)+2^{-8n_\chi}\left(\chi_d+2^{8n_\chi-1}+1/2\right),
\end{equation}
where $[\ ]$ is the operator to take the integer part. $n_\chi\in\{1,2\}$ is the number of bytes used for each integer, $x_d$ and $\chi_d$ are floating and integer versions of the coordinate. The velocity counterpart of them are $n_\nu=1,2$, $v_d$ and $\nu_d$. 
%$n_c=1,2,...$ is the coarse grid index, given by the ordering of particles and a integer based particle density field (see \ref{sss.spatial_decomposition}).
The position resolution for $n_\chi$-byte integer, ``x$n_\chi$'', is $2^{-8n_\chi}L/N_c$.

@@

\subsection{Particle velocity storage}\label{ss.velocity}
Similarly, actual velocity in $d$th dimension $v_d$ is decomposed into an averaged velocity field on the same coarse grid $v_c$ and a residual $\Delta v$ relative to this field: 
\begin{equation}
	v_d=v_c+\Delta v.
\end{equation}
$v_c$ is always recorded and kept updated, and should not occupy considerable memory. We then divide velocity space $\Delta v$ into uneven bins, and use a $n_\nu$-byte integer to indicate which $\Delta v$ bin the particle is located.

@@

\begin{figure}
\centering
  \includegraphics[width=0.9\linewidth]{f2}
 \caption{Spacial decomposition in {\tt CUBE} in a 2D analogy. In this example, there are 2 images per dimension ($M_g=2$), and 2 tiles per image per dimension ($M_t=2$). The orange boxes show the overlapped extended($V_e$)=physical($V_p$)+buffer($V_b$) tile regions. One physical region is indicated with green.}
\label{fig.tile}
\end{figure}

The simulation results are very insensitive if we manually tune the variance of the model $\sigma_\Delta$ within an order of magnitude. However, in $n_\nu=1$ case, the method of using uneven bins gets much better results than simply using equal bins between minimum and maximum values $[\min(\Delta v),\max(\Delta v)]$. So, one can safely use a standard $\Lambda$CDM (cold dark matter with a cosmological constant as dark energy) for slightly different cosmological models, in equation (\ref{eq.vdisp}). In {\tt CUBE}, the velocity conversion takes the formula
\begin{equation}\label{eq.nu}
	\nu_d=\left\lfloor(2^{8n_\nu}-1)\pi^{-1}\tan^{-1}\left((v_d-v_c)\sqrt{\pi/2\sigma_\Delta^2}\right)\right\rfloor,
\end{equation}
\begin{equation}\label{eq.v}
	v_d=v_c+\tan\left(\frac{\pi\nu_d}{2^{8n_\nu}-1}\right)\sqrt{2\sigma_\Delta^2/\pi},
\end{equation}
where $\lfloor\ \rfloor$ is the operator to take the nearest integer. Tangent functions are convenient and computing-efficient. Compared to error functions used in Gaussian case, they take the same variance at $\nu_d=0$ but resolve high velocities relatively better. All possible choice of conversion formulae and $\sigma_\Delta$ are unbiased in the conversion Equations (\ref{eq.nu},\ref{eq.v}), however a proper choices optimize the velocity space sampling and can result in more precise results.

@@

\subsection{Code overview}\label{ss.overview}
{\tt CUBE} uses a 2-level PM force calculation. In order to apply the integer based format to the $N$-body simulation, substantial structural changes need to be done. {\tt CUBE} is written in Coarray Fortran, where Coarray features replace MPI (Message Passing Interface) communications between computation nodes/images\footnote{Images are the concept of computing nodes or MPI tasks in Coarray Fortran. We use this terminology in this paper.}. The algorithm is described in this language.

\subsubsection{Spacial decomposition}\label{sss.spatial_decomposition}
{\tt CUBE} decomposes the global simulation volume into $M_g^3$ cubic sub-volumes with $N_c$ coarse grids or $N_f=RN_c$ fine grids per side. The fine-mesh is usually $R=4$ times finer than the coarse mesh. Each of these sub-volumes is assigned to a coarray {\it image}. Inside of an image, the sub-volume is further decomposed into $M_t^3$ cubic {\it tiles} (defined as $V_p$, or {\it physical} region) with $N_c/(M_g M_t)$ coarse grids per side. Each $V_p$ is surrounded by a {\it buffer} region $V_b$ which is $N_b$ coarse cells thick. We define extended tile region $V_e\equiv V_p+V_b$. $V_e$ is designed for two reasons: (1) if the short fine-mesh force ${\bs F}_f$ has a cut-off $N_b$, and is computed by $\rho_c$ whose domain $D(\rho_f)=V_e$, then ${\bs F}_f$ in $V_p$ is guaranteed to be correct, i.e.,
\begin{eqnarray}\label{eq.cond1}
\nonumber	{\bs F}_f(r>N_b)&=&0 \wedge D(\rho_c)=V_e\\
	&\Rightarrow& \{{\bs F}_f(\rho_c) |{\bs F}_f\in V_p \} {\rm \ is\ correct},
\end{eqnarray}
and (2) to collect all particles being able to travel to $V_p$, i.e.,
\begin{equation}\label{eq.cond2}
	{\bs v}\,{\rm d}t \leq N_b \Rightarrow \{ {\bs x} | {\bs x}+{\bs v}\,{\rm d}t\in V_p\} \in V_e.
\end{equation}
Figure \ref{fig.tile} shows the spacial decomposition in a 2-dimensional analogy, with $M_g=2$ and $M_t=2$.

According to this spatial decomposition, and as discussed in the last two subsections, we declare $\{\rho_c,v_c,\chi_d,\nu_d\}$ by using Fortran language:
\begin{eqnarray*}
	&&{\tt integer(1)}\ \ \, \rho_c(N_e,N_e,N_e,M_t,M_t,M_t)[M_g,M_g,*]\\
	&&{\tt real(4)}\ \ \ \ \ \ \ \, v_c(N_e,N_e,N_e,M_t,M_t,M_t)[M_g,M_g,*]\\
	&&{\tt integer(}n_\chi{\tt )}\, \chi_d(3,P_{\rm max})[M_g,M_g,*]\\
	&&{\tt integer(}n_\nu{\tt )}\ \nu_d(3,P_{\rm max})[M_g,M_g,*]
\end{eqnarray*}
where $N_e=N_t+2N_b$ covers the buffer region on both sides, $M_t$ is the tile dimemsions, and $M_g$ is the image {\it codimensions}\footnote{Coarray Fortran concept. Codimensions can do communications between images.}. We denote actual number of particles in a given image $P_{\rm local}$, and $P_{\rm max}>P_{\rm local}$ is a value (discussed in Section \ref{ss.memory}) large enough to store particles in $V_e$. $\chi_d$ and $\nu_d$
%(equivalent to $\chi_d$ and $\nu_d$ in \ref{ss.position} and \ref{ss.velocity} respectively)
are required to be sorted according to the same memory layout of $\rho_c$, such that $n_c$ and $x_d$ can be obtained by equation (\ref{eq.x}). 
$\{\rho_c,v_c,\chi_d,\nu_d\}$ provides a complete information of positions and velocities of particles, and we call it a checkpoint.

An additional particle-ID (PID) array, $I_P$, can also be declared to differentiate particle types (e.g. CDM and neutrino particles) or differentiate every particle, by using $n_I$-integer per particle. If PID is turned on, the array $I_P$ is also included in the checkpoints, and the ordering of $I_P$ is same as $\chi_d$ and $\nu_d$.

\begin{figure}[t]
{\tt program Initial\_Condition\_Generator\_for\_CUBE\\
\indent \ \ calculate $\Phi$ in Fourier space\\
\indent \ \ do (each tile)\\
\indent \ \ \ \ do (each particle at ${\bs q}\in V_e$)\\
\indent \ \ \ \ \ \ ${\bs \Psi}({\bs q})=\nabla\Phi({\bs q})$\\
\indent \ \ \ \ \ \ $\{{\bs q},{\bs\Psi({\bs q})}\}\rightarrow\{{\bs x},{\bs v}\}$\\
\indent \ \ \ \ \ \ $\{{\bs x},{\bs v}\}\rightarrow \{\rho_c,v_c\}$\\
\indent \ \ \ \     enddo\\
\indent \ \ \ \ do (each particle at ${\bs q}\in V_e$)\\
\indent \ \ \ \ \ \ ${\bs \Psi}({\bs q})=\nabla\Phi({\bs q})$\\
\indent \ \ \ \ \ \ $\{{\bs q},{\bs\Psi({\bs q})}\}\rightarrow\{{\bs x},{\bs v}\}$\\
\indent \ \ \ \ \ \ calculate particle's index $i$ according to $\rho_c$\\
\indent \ \ \ \ \ \ ${\bs x}\rightarrow\chi_d(:,i);\,\{{\bs v},v_c\}\rightarrow\nu_d(:,i)$\\
\indent \ \ \ \ \ \ if (PID\_flag) create $I_P(i)$\\
\indent \ \ \ \ enddo\\
\indent \ \ \ \ do (each coarse grid $\in V_e$)\\
\indent \ \ \ \ \ \ delete particles $\in V_b$\\
\indent \ \ \ \ enddo\\
\indent \ \ \ \ write $\{\rho_c,v_c,\chi_d,\nu_d\}$ to disk\\
\indent \ \ \ \ sum up $P_{\rm local}$\\
\indent \ \ enddo\\
\indent \ \ sync all\\
\indent \ \ sum up $P_{\rm global}$\\
end\\}
\caption{Pseudocode for the initial condition generator.}
\label{fig.ic}
\end{figure}

@@

The first step is the calculation of the displacement potential $\Phi$. At an initial redshift $z_i$, we generate a linear density fluctuation $\delta_L({\bs q})$ on Lagrangian grid ${\bs q}$ by multiplying a Gaussian random field with the transfer function $T(k)$ (given by the assumed cosmological model) in Fourier space. Then we solve for the potential ${\Phi({\bs q}, z_i)}$ of a curl-less displacement field ${\bs\Psi({\bs q})}$ by Poisson equation $-\nabla^2\Phi=\delta_L$ in Fourier space.

The second step to generate particles and displace them by Zel'dovich approximation (ZA) \citep{1970A&A.....5...84Z}, where the displacement field is obtained by differentiating $\Phi$, ${\bs \Psi}({\bs q})=\nabla\Phi({\bs q})$ in real space. This step is done on $V_e$ of each tile. Similar to Equation (\ref{eq.cond2}), we have
\begin{equation}\label{eq.cond3}
	\max(\Psi_d) \leq N_b \Rightarrow \{ {\bs q} | {\bs q}+{\bs \Psi}({\bs q})\in V_p\} \in V_e.
\end{equation}
We iterate twice over particles' ${\bs q}$ in $V_e$. The first iteration calculates particles' ${\bs x}$ and ${\bs v}$ by ZA and obtains $\rho_c$ and $v_c$ on the coarse grid. The second iteration, ${\bs x}$ and ${\bs v}$ is calculated again and are converted to $\chi_d$ and $\nu_d$ by Equations (\ref{eq.chi},\ref{eq.nu}) and place them in a certain order according to $\rho_c$. Lastly, we delete particles in $V_b$, and resort the ones in $V_p$ and write $\{\rho_c,v_c,\chi_d,\nu_d,I_P({\rm optional})\}$ of this tile to disk. The above is similar to {\tt update\_xp} of Section \ref{sss.algorithm}. If PIDs are needed, they are also generated here.
After working on all tiles, we sum up $P_{\rm local}$ and $P_{\rm global}$. We summarize the above steps into a pseudocode in Figure \ref{fig.ic}. During this step, the only major memory usage is $\Phi$ on the fine mesh. If the number of particles per fine grid $P_f=1$, the memory consumption of this in-place FFT is 4 bpp.

@@

\subsubsection{Algorithm}\label{sss.algorithm}
Figure \ref{fig.code} shows the overall structure of the main code.

{\tt initialize} -- creates fine-mesh and coarse mesh FFT plans, and read in configuration files telling the program at which redshifts we need to do checkpoints, halofinds, or stop the simulation. Force kernels $K_c$, $K_f$ are also computed or loaded.

{\tt read\_particles} -- from disk, reads in checkpoint $\{\rho_c,v_c,\chi_d,\nu_d,I_P({\rm optional})\}$ for each image. Because they exist only in $V_p$ of every tile, they are {\it disjoint}; and they provide {\it complete} information of the whole simulation volume -- we call it ``{\it disjoint state}''. In this state, $\rho_c$, $v_c$'s values in buffer regions, and $\chi_d(:,P_{\rm local}+1:)$ and $\nu_d(:,P_{\rm local}+1:)$ are 0's. Because $I_P$ is generated and manipulated together with $\nu_d$, so we do not explicitly mention $I_P$ in the followings.

{\tt buffer\_density}, {\tt buffer\_x} and {\tt buffer\_v} -- converts ``disjoint'' state to ``{\it buffered state}''. In {\tt buffer\_density}, $V_b$ regions of $\rho_c$ is synchronized between tiles and images. By {\tt buffer\_x}, $\chi_d$ is updated to contain common, buffered particles, and they are sorted according to the buffered $\rho_c$. {\tt buffer\_v} deals with $\nu_d$ in a same manner.

{\tt timestep} --
Runge-Kutta 2 method is used in the time integration. i.e., for $n$ time steps, we update positions ($D$=drift) and velocities ($K$=kick) at interlaced half time-steps by operator splitting; the operation $(DKKD)^n$ is 2nd order accurate. The actual simulation applies varied time steps by {\tt timestep}, where a time increment ${\rm d}t$ is constrained by particles' maximum velocities, accelerations, cosmic expansion and any other desired conditions.
